High-level cognition during story listening is reflected in high-order dynamic correlations in neural activity patterns

### 2019-09-05 paper/main.tex
updated patterns figure, inputing notes to textt

\title{High-level cognition is supported by at least second order
  dynamic correlations in neural activity patterns}

@@

\begin{abstract}
  Our thoughts arise from coordinated patterns of interactions between
  brain structures that change with our ongoing experiences.
  High-order dynamic correlations in brain activity patterns reflect
  different subgraphs of the brain's connectome that display
  homologous lower-level dynamic correlations.  We tested the
  hypothesis that high-level cognition is supported by high-order
  dynamic correlations in brain activity patterns.  We developed an
  approach to estimating high-order dynamic correlations in timeseries
  data, and we applied the approach to neuroimaging data collected as
  human participants either listened to a ten-minute story or a
  temporally scrambled version of the story, or underwent a resting
  state scan.  We trained across-participants pattern classifiers to
  decode (in held-out data) when in the session each activity snapshot
  was collected.  We found that classifiers trained to decode from
  high-order dynamic correlations yielded better performance on data
  collected as participants listened to the (unscrambled) story.  By
  contrast, classifiers trained to decode data from scrambled versions
  of the story or during the resting state scan yielded the best
  performance when they were trained using first-order dynamic
  correlations or raw activity patterns.  We suggest that as our
  thoughts become more complex, they are supported by higher-order
  patterns of dynamic network interactions throughout the brain.
\end{abstract}

@@

To gain insights into the above aspects of the neural code, we
developed a computational framework for estimating dynamic high-order
correlations in timeseries data. This framework provides an important
advance, in that it enables us to examine patterns in higher-order
correlations that are computationally intractable to estimate via
conventional methods.  Given a multivariate timeseries, our framework
provides timepoint-by-timepoint estimates of the first-order
correlations, second-order correlations, and so on (up to tenth-order
correlations in this manuscript).  Our approach combines a
kernel-based method for computing dynamic correlations in timeseries
data with a dimensionality reduction step (Fig.~\ref{fig:patterns}b)
that projects the resulting dynamic correlations into a
low-dimensional space.  We explored two dimensionality reduction
approaches: principle components analysis~\citep[PCA;][]{Pear01},
which preserves an approximately invertible transformation back to the
original data~\citep[e.g., this follows related approaches taken
by][]{McInJirs19, TokeSomm19, GonzEtal19}; and a second non-invertible
algorithm that explored patterns in eigenvector
centrality~\citep{Land95}.  This latter approach characterizes
correlations between each feature dimension's relative
\textit{position} in the network in favor of the specific activity
histories of different features~\citep[also see][]{BetzEtal19,
  SizeEtal18, ReimEtal17}.

We validated our approach using synthetic data where the underlying
correlations were known.  We then applied our framework to a
neuroimaging dataset collected as participants
listened to either an audio recording of a ten-minute story or a
temporally scrambled version of the story, or underwent a resting
state scan~\citep{SimoEtal16}.  We used a subset of the data to train
across-participant classifiers to decode listening times using a blend
of neural features (comprising neural activity patterns, as well as
different orders of correlations between those patterns that were
inferred using our computational framework).  We found that both the
PCA-based and eigenvector centrality-based approaches yielded neural
patterns that could be used to decode accurately.  Both approaches
also yielded the best decoding accuracy for data collected during
(intact) story listening when high-order (PCA: second-order;
eigenvector centrality: fourth-order) dynamic correlation patterns
were included as features.  When we trained classifiers on the
scrambled stories or resting state data, only lower-order dynamic
patterns were informative to the decoders.  Taken together, our
results indicate that high-level cognition is supported by high-order
dynamic patterns of communication between brain structures.

\section*{Methods}
Our general approach to comprises four general steps
(Fig.~\ref{fig:methods}).  First, we derive a kernel-based approach to
computing dynamic pairwise correlations in a $T$ (timepoints) by $K$
(features) multivariate timeseries, $\mathbf{X}_0$.  This yields a $T$
by $\mathcal{O}(K^2)$ matrix of dynamic correlations, $\mathbf{Y}_1$,
where each row comprises the upper triangle of the correlation matrix
at a single timepoint, reshaped into a row vector (this reshaped
vector is $(\frac{K^2 - K}{2})$-dimensional).  Second, we apply a
dimensionality reduction step to project the matrix of dynamic
correlations back onto a $K$-dimensional space.  This yields a $T$ by
$K$ matrix, $\mathbf{X}_1$, that reflects an approximation of the
dynamic correlations reflected in the original data.  Third, we use
repeated applications of the kernel-based dynamic correlation step to
$\mathbf{X}_n$ and the dimensionality reduction step to the resulting
$\mathbf{Y}_{n+1}$ to estimate high-order dynamic correlations.  Each
application of these steps to a $T$ by $K$ time series $\mathbf{X}_n$
yields a $T$ by $K$ matrix, $\mathbf{X}_{n+1}$, that reflects the
dynamic correlations between the columns of $\mathbf{X}_n$.  In this
way, we refer to $n$ as the \textit{order} of the timeseries, where
$\mathbf{X}_0$ (order 0) denotes the original data and $\mathbf{X}_n$
denotes $n^\mathrm{th}$-order dynamic correlations between the columns
of $\mathbf{X}_0$.  Finally, we use a cross-validation--based decoding
approach to evaluate how well information contained in a given order
(or weighted mixture of orders) may be used to decode relevant
cognitive states.  If including a given $\mathbf{X}_n$ in the feature
set yields higher classification accuracy on held-out data, we
interpret this as evidence that the given cognitive states are
reflected in patterns of $n^\mathrm{th}$-order correlations.  All of
the code used to produce the figures and results in this manuscript,
along with links to the corresponding datasets, may be found at
\href{https://github.com/ContextLab/timecorr-paper}{github.com/ContextLab/timecorr-paper}.  In addition, we have
released a Python toolbox for computing dynamic high-order
correlations in timeseries data; our toolbox may be found at
\href{https://timecorr.readthedocs.io/}{timecorr.readthedocs.io}. \textbf{JRM NOTE: CHECK LINK}

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{figs/methods_fig}
  \caption{\textbf{Estimating dynamic high-order correlations.}  Given
    a $T$ by $K$ matrix of multivariate timeseries data,
    $\mathbf{Y}_n$ (where $n \in \mathbb{N}, n \geq 0$), we use
    Equation~\ref{eqn:timecorr} to compute a timeseries of
    $K$ by $K$ correlation matrices, $\mathbf{Y}_{n+1}$.  We then
    approximate $\mathbf{Y}_{n+1}$ with the $T$ by $K$ matrix
    $\mathbf{X}_{n+1}$.  This process may be repeated to scalably estimate
    iteratively higher-order correlations in the data.  Note that the
    transposes of $\mathbf{Y}_n$ and $\mathbf{X}_{n+1}$ are displayed
    in the figure for compactness.
  \label{fig:methods}}
\end{figure}

\subsection*{Kernel-based approach for computing dynamic correlations}
Given a matrix of observations, we can compute the (static)
Pearson's correlation between any pair of columns, $\mathbf{X}(\cdot, i)$ and
$\mathbf{X}(\cdot, j)$ using~\citep{Pear01}:
\begin{align}
  \mathrm{corr}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)) &=
                                                              \frac{\sum_{t=1}^T
                                                              \left(\mathbf{X}(t,
                                                              i)
                                                              -
                                                              \bar{\mathbf{X}(\cdot,
                                                              i)}\right)
                                                              \left(\mathbf{X}(t,
                                                              j)
                                                              -
                                                              \bar{\mathbf{X}(\cdot, j)}\right)}{\sqrt{\sum_{t=1}^T
                                                              \sigma^2_{\mathbf{X}(\cdot, i)} 
                                                              \sigma^2_{\mathbf{X}(\cdot, j)}}},~\mathrm{where}\\\label{eqn:corr}
  \bar{\mathbf{X}(\cdot, k)} &= \sum_{t=1}^T
                       \mathbf{X}(t, k),~\mathrm{and}\\
  \sigma^2_{\mathbf{X}(\cdot, k)} &= \sum_{t=1}^T \left( \mathbf{X}(t, k) -
                            \bar{\mathbf{X}(\cdot, k)} \right)^2 
\end{align}
We can generalize this formula to compute time-varying correlations by
incorporating a \textit{kernel function} that takes a time $t$ as
input, and returns how much the observed data at each timepoint
$\tau \in \left[ -\infty, \infty \right]$ contributes to the estimated instantaneous
correlation at time $t$~\citep[Fig.~\ref{fig:kernels}; also see][for a
similar approach]{AlleEtal12b}.
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/kernels}
  \caption{\textbf{Examples of kernel functions.} Each panel
    displays per-timepoint weights at $t = 50$, evaluated for 100
    timepoints ($\tau \in \left[1, ..., 100\right]$).  \textbf{a. Uniform
      kernel.} The weights are timepoint-invariant; observations at
    all timepoints are weighted equally, and do not change as a
    function of $\tau$.  This is a special case of weight function that
    reduces dynamic correlations to static correlations.
    \textbf{b. Dirac $\delta$ kernel.} Only the observation at
    timepoint $t$ is given a non-zero weight (of 1).
    \textbf{c. Gaussian kernels.} Each kernel's weights fall off
    in time according to a Gaussian probability density function
    centered on time $t$.  Weights derived using several different
    example width parameters ($\sigma^2$) are displayed.
    \textbf{d. Laplace kernels.}  Each kernel's weights fall off
    in time according to a Laplace probability density function
    centered on time $t$.  Weights derived using several different
    example width parameters ($b$) are displayed.  \textbf{e. Mexican
      hat (Ricker wavelet) kernels.}  Each kernel's weights fall
    off in time according to a Ricker wavelet centered on time $t$.  This
    function highlights the \textit{contrasts} between local versus
    surrounding activity patterns in estimating dynamic
    correlations. Weights derived using several different example
    width parameters ($\sigma$) are displayed.}
  \label{fig:kernels}
\end{figure}

Given a kernel function $\kappa_t(\cdot)$ for timepoint $t$,
evaluated at timepoints $\tau \in \left[ 1, ..., T \right]$, we
can update the static correlation formula in Equation~\ref{eqn:corr}
to estimate the \textit{instantaneous correlation} at timepoint $t$:
\begin{align}
  \mathrm{timecorr}_{\kappa_t}\left(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)\right) &= \frac{\sum_{\tau=1}^T \left( \mathbf{X}(\tau, i) -
                                       \widetilde{\mathbf{X}}_{\kappa_t}(i) \right)
                                 \left( \mathbf{X}(\tau, j) -
                                        \widetilde{\mathbf{X}}_{\kappa_t}(j)\right)}
              {\sqrt{\sum_{\tau=1}^T
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot,
                                                                                        i))
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, j))}},~\mathrm{where}\\\label{eqn:timecorr}
  \widetilde{\mathbf{X}}_{\kappa_t}(t, k) &= \sum_{\tau=1}^T
                       \kappa_t(\tau, k)\mathbf{X}(\tau, k),\\
  \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, k)) &= \sum_{\tau=1}^T
                                                  \left(
                                                  \mathbf{X}(\tau, k) -
                            \widetilde{\mathbf{X}}_{\kappa_t}(t, k) \right)^2.
\end{align}
Here
$\mathrm{timecorr}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j),
\kappa_t)$ reflects the correlation at time $t$ between columns $i$
and $j$ of $\mathbf{X}$, estimated using the kernel $\kappa_t$.  We
evaluate Equation~\ref{eqn:timecorr} in turn each pair of columns in
$\mathbf{X}$ and for kernels centered on each timepoint in the
timeseries, respectively, to obtain a $T$ by $K$ by $K$ timeseries of
dynamic correlations, $\mathbf{Y}$.  For convenience, we then reshape
the upper triangles of each timepoint's correlation matrix into a row
vector to obtain an equivalent $T$ by $\frac{K^2 - K}{2}$ matrix.

\subsubsection*{Dynamic inter-subject functional connectivity (DISFC)}
Equation~\ref{eqn:timecorr} provides a means of taking a single
observation matrix, $\mathbf{X}_n$ and estimating the dynamic
correlations from moment to moment, $\mathbf{Y}_{n+1}$.  Suppose that
one has access to a set of multiple observation matrices that reflect
the same phenomenon.  For example, one might collect neuroimaging data
from several experimental participants, as each participant performs
the same task (or sequence of tasks).  Let $\mathbf{X}_n^1$,
$\mathbf{X}_n^2$, ..., $\mathbf{X}_n^P$ reflect the $T$ by $K$
observation matrices ($n = 0$) or reduced correlation matrices
($n > 0$) for each of $P$ participants in an experiment.  We can use
\textit{inter-subject functional connectivity}~\citep[ISFC;
][]{SimoEtal16} to compute the degree of stimulus-driven correlations
reflected in the multi-participant dataset at a given timepoint $t$
using:
\begin{align}
\bar{\mathbf{C}}(t) = M\left(R\left(\frac{1}{2P} \sum_{p=1}^P
  Z\left(Y_n^p(t)\right)^\mathrm{T} + Z\left(Y_n^p(t)\right)\right)\right),\label{eqn:disfc}
\end{align}
where $M$ extracts and vectorizes the diagonal and upper triangle of a symmetric
matrix, $Z$ is the Fisher $z$-transformation~\citep{Zar10}:
\begin{align}
Z(r) = \frac{\log(1+r) - \log(1-r)}{2}
\end{align}
$R$ is the inverse of $Z$:
\begin{align}
R(z) = \frac{\exp(2z - 1)}{\exp(2z + 1)},
\end{align}
and $\mathbf{Y}_n^p(t)$ denotes the correlation matrix
(Eqn.~\ref{eqn:corr}) between each column of $\mathbf{X}_n^p$ and each
column of the average $\mathbf{X}_n$ from all \textit{other}
participants, $\bar{\mathbf{X}}_n^{ \backslash p}$:
\begin{align}
  \bar{\mathbf{X}}_n^{ \backslash p} = R\left(\frac{1}{P-1}\sum_{q \in
  \backslash p} Z\left( \mathbf{X}_n^q \right) \right),
\end{align}
where $\backslash p$ denotes the set of all participants other than
participant $p$. In this way, the $T$ by $\frac{K^2 - K}{2}$ DISFC
matrix $\bar{\mathbf{C}}$ provides a time-varying extension of the ISFC
approach developed by \cite{SimoEtal16}.

@@

Given a $T$ by $\frac{K^2 - K}{2}$ matrix of dynamic correlations,
$\mathbf{Y}_n$, we propose two general approaches to computing a $T$
by $K$ low-dimensional representation of these correlations,
$\mathbf{X}_n$.  The first approach uses dimensionality reduction
algorithms to project $\mathbf{Y}_n$ onto a $K$-dimensional space.
The second approach uses graph measures to characterize the
relative positions of each feature ($k \in \left[1, ..., K \right]$)
in the network defined by the correlation matrix at each timepoint.

@@

The modern library of dimensionality reduction algorithms include
Principal Components Analysis~\citep[PCA; ][]{Pear01}, Probabilistic
PCA~\citep[PPCA; ][]{TippBish99}, Exploratory Factor
Analysis~\citep[EFA; ][]{Spea04}, Independent Components
Analysis~\citep[ICA; ][]{JuttHera91, ComoEtal91}, $t$-Stochastic
Neighbor Embedding~\citep[$t$-SNE; ][]{MaatHint08}, Uniform Manifold
Approximation and Projection~\citep[UMAP; ][]{McInHeal18},
non-negative matrix factorization~\citep[NMF; ][]{LeeSeun99},
Topographic Factor Analysis (TFA)~\cite{MannEtal14b}, Hierarchical
Topographic Factor analysis (HTFA)~\cite{MannEtal18}, Topographic
Latent Source Analysis (TLSA)~\cite{GersEtal11}, Dictionary
learning~\citep{MairEtal09a, MairEtal09b}, deep
auto-encoders~\citep{HintSala06}, among others.  While complete
characterizations of each of these algorithms is beyond the scope of
the present manuscript, the general intuition driving these approaches
is to compute the $T$ by $I$ matrix, $\mathbf{X}$, that is closest
to the original $T$ by $J$ matrix, $\mathbf{Y}$, where (typically)
$I \ll J$.  The different approaches place different constraints on
what properties $\mathbf{X}$ must satisfy and which aspects of
the data are compared (and how) to characterize the match between
$\mathbf{X}$ and $\mathbf{Y}$.

Applying dimensionality reduction algorithms to $\mathbf{Y}$ yields a
$\mathbf{X}$ whose columns reflect weighted combinations (or nonlinear
transformations) of the original columns of $\mathbf{Y}$.  This has
two main consequences.  First, with each repeated dimensionality
reduction, the resulting $\mathbf{X}_n$ has lower and lower fidelity
(with respect to what the ``true'' $\mathbf{Y}_n$ might have looked
like without using dimensionality reduction to maintain scalability).
In other words, computing $\mathbf{X}_n$ is a lossy operation.
Second, whereas each columns of $\mathbf{Y}_n$ may always be mapped
directly onto specific pairs of columns of $\mathbf{Y}_{n-1}$, the
columns of $\mathbf{X}_n$ reflect weighted combinations and/or
nonlinear transformations of the columns of $\mathbf{Y}_n$.  Many
dimensionality reduction algorithms are invertible (or approximately
invertible).  However, attempting to map a given $\mathbf{X}_n$ back
onto the original feature space of $\mathbf{Y}_0$ will usually require
$\mathcal{O}(TK^{2n})$ space and therefore quickly becomes intractable
as $n$ or $K$ grow large.

>>>>

\title{High-level cognition during story listening is reflected in
  high-order dynamic correlations in neural activity patterns}

@@

\begin{abstract}
  Our thoughts arise from coordinated patterns of interactions between
  brain structures that change with our ongoing experiences.
  High-order dynamic correlations in brain activity patterns reflect
  different subgraphs of the brain's connectome that display
  homologous lower-level dynamic correlations.  We tested the
  hypothesis that high-level cognition is supported by high-order
  dynamic correlations in brain activity patterns.  We developed an
  approach to estimating high-order dynamic correlations in timeseries
  data, and we applied the approach to neuroimaging data collected as
  human participants either listened to a ten-minute story, listened
  to a temporally scrambled version of the story, or underwent a
  resting state scan.  We trained across-participants pattern
  classifiers to decode (in held-out data) when in the session each
  neural activity snapshot was collected.  We found that classifiers
  trained to decode from high-order dynamic correlations yielded the
  best performance on data collected as participants listened to the
  (unscrambled) story.  By contrast, classifiers trained to decode
  data from scrambled versions of the story or during the resting
  state scan yielded the best performance when they were trained using
  first-order dynamic correlations or non-correlational activity patterns.  We
  suggest that as our thoughts become more complex, they are supported
  by higher-order patterns of dynamic network interactions throughout
  the brain.
\end{abstract}

@@

To gain insights into the above aspects of the neural code, we
developed a computational framework for estimating dynamic high-order
correlations in timeseries data. This framework provides an important
advance, in that it enables us to examine patterns in higher-order
correlations that are computationally intractable to estimate via
conventional methods.  Given a multivariate timeseries, our framework
provides timepoint-by-timepoint estimates of the first-order
correlations, second-order correlations, and so on.  Our approach
combines a kernel-based method for computing dynamic correlations in
timeseries data with a dimensionality reduction step
(Fig.~\ref{fig:patterns}b) that projects the resulting dynamic
correlations into a low-dimensional space.  We explored two
dimensionality reduction approaches: principle components
analysis~\citep[PCA;][]{Pear01}, which preserves an approximately
invertible transformation back to the original data~\citep[e.g., this
follows related approaches taken by][]{McInJirs19, TokeSomm19,
  GonzEtal19}; and a second non-invertible algorithm that explored
patterns in eigenvector centrality~\citep{Land95}.  This latter
approach characterizes correlations between each feature dimension's
relative \textit{position} in the network in favor of the specific
activity histories of different features~\citep[also
see][]{BetzEtal19, SizeEtal18, ReimEtal17}.

We validated our approach using synthetic data where the underlying
correlations were known.  We then applied our framework to a
neuroimaging dataset collected as participants listened to either an
audio recording of a ten-minute story or a temporally scrambled
version of the story, or underwent a resting state
scan~\citep{SimoEtal16}.  We used a subset of the data to train
across-participant classifiers to decode listening times using a blend
of neural features (comprising neural activity patterns, as well as
different orders of dynamic correlations between those patterns that
were inferred using our computational framework).  We found that both
the PCA-based and eigenvector centrality-based approaches yielded
neural patterns that could be used to decode accurately (well above
chance).  Both approaches also yielded the best decoding accuracy for
data collected during (intact) story listening when high-order (PCA:
second-order; eigenvector centrality: fourth-order) dynamic
correlation patterns were included as features.  When we trained
classifiers on the scrambled stories or resting state data, only
lower-order dynamic patterns were informative to the decoders.  Taken
together, our results indicate that high-level cognition is supported
by high-order dynamic patterns of communication between brain
structures.

\section*{Methods}
Our general approach to comprises four general steps
(Fig.~\ref{fig:methods}).  First, we derive a kernel-based approach to
computing dynamic pairwise correlations in a $T$ (timepoints) by $K$
(features) multivariate timeseries, $\mathbf{X}_0$.  This yields a $T$
by $\mathcal{O}(K^2)$ matrix of dynamic correlations, $\mathbf{Y}_1$,
where each row comprises the upper triangle of the correlation matrix
at a single timepoint, reshaped into a row vector (this reshaped
vector is $(\frac{K^2 - K}{2})$-dimensional).  Second, we apply a
dimensionality reduction step to project the matrix of dynamic
correlations back onto a $K$-dimensional space.  This yields a $T$ by
$K$ matrix, $\mathbf{X}_1$, that reflects an approximation of the
dynamic correlations reflected in the original data.  Third, we use
repeated applications of the kernel-based dynamic correlation step to
$\mathbf{X}_n$ and the dimensionality reduction step to the resulting
$\mathbf{Y}_{n+1}$ to estimate high-order dynamic correlations.  Each
application of these steps to a $T$ by $K$ time series $\mathbf{X}_n$
yields a $T$ by $K$ matrix, $\mathbf{X}_{n+1}$, that reflects the
dynamic correlations between the columns of $\mathbf{X}_n$.  In this
way, we refer to $n$ as the \textit{order} of the timeseries, where
$\mathbf{X}_0$ (order 0) denotes the original data and $\mathbf{X}_n$
denotes (approximated) $n^\mathrm{th}$-order dynamic correlations
between the columns of $\mathbf{X}_0$.  Finally, we use a
cross-validation--based decoding approach to evaluate how well
information contained in a given order (or weighted mixture of orders)
may be used to decode relevant cognitive states.  If including a given
$\mathbf{X}_n$ in the feature set yields higher classification
accuracy on held-out data, we interpret this as evidence that the
given cognitive states are reflected in patterns of
$n^\mathrm{th}$-order correlations.  All of the code used to produce
the figures and results in this manuscript, along with links to the
corresponding datasets, may be found at
\href{https://github.com/ContextLab/timecorr-paper}{github.com/ContextLab/timecorr-paper}.
In addition, we have released a Python toolbox for computing dynamic
high-order correlations in timeseries data; our toolbox may be found
at
\href{https://timecorr.readthedocs.io/}{timecorr.readthedocs.io}. \textbf{JRM
  NOTE: CHECK LINK}

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{figs/methods_fig}
  \caption{\textbf{Estimating dynamic high-order correlations.}  Given
    a $T$ by $K$ matrix of multivariate timeseries data,
    $\mathbf{X}_n$ (where $n \in \mathbb{N}, n \geq 0$), we use
    Equation~\ref{eqn:timecorr} to compute a timeseries of
    $K$ by $K$ correlation matrices, $\mathbf{Y}_{n+1}$.  We then
    approximate $\mathbf{Y}_{n+1}$ with the $T$ by $K$ matrix
    $\mathbf{X}_{n+1}$.  This process may be repeated to scalably estimate
    iteratively higher-order correlations in the data.  Note that the
    transposes of $\mathbf{X}_n$ and $\mathbf{X}_{n+1}$ are displayed
    in the figure for compactness.
  \label{fig:methods}}
\end{figure}

\subsection*{Kernel-based approach for computing dynamic correlations}
Given a $T$ by $K$ matrix of observations, $\mathbf{X}$, we can compute the (static)
Pearson's correlation between any pair of columns, $\mathbf{X}(\cdot, i)$ and
$\mathbf{X}(\cdot, j)$ using~\citep{Pear01}:
\begin{align}
  \mathrm{corr}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)) &=
                                                              \frac{\sum_{t=1}^T
                                                              \left(\bar{\mathbf{X}}(t,
                                                              i)
                                                              -
                                                              \mathbf{X}(\cdot,
                                                              i)\right)
                                                              \left(\mathbf{X}(t,
                                                              j)
                                                              -
                                                              \bar{\mathbf{X}}(\cdot, j)\right)}{\sqrt{\sum_{t=1}^T
                                                              \sigma^2_{\mathbf{X}(\cdot, i)} 
                                                              \sigma^2_{\mathbf{X}(\cdot, j)}}},~\mathrm{where}\\\label{eqn:corr}
  \bar{\mathbf{X}}(\cdot, k) &= \frac{1}{T}\sum_{t=1}^T
                       \mathbf{X}(t, k),~\mathrm{and}\\
  \sigma^2_{\mathbf{X}(\cdot, k)} &= \frac{1}{T}\sum_{t=1}^T \left( \mathbf{X}(t, k) -
                            \bar{\mathbf{X}}(\cdot, k) \right)^2 
\end{align}
We can generalize this formula to compute time-varying correlations by
incorporating a \textit{kernel function} that takes a time $t$ as
input, and returns how much the observed data at each timepoint
$\tau \in \left[ -\infty, \infty \right]$ contributes to the estimated instantaneous
correlation at time $t$~\citep[Fig.~\ref{fig:kernels}; also see][for a
similar approach]{AlleEtal12b}.
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/kernels}
  \caption{\textbf{Examples of kernel functions.} Each panel
    displays per-timepoint weights at $t = 50$, evaluated for 100
    timepoints ($\tau \in \left[1, ..., 100\right]$).  \textbf{a. Uniform
      kernel.} The weights are timepoint-invariant; observations at
    all timepoints are weighted equally, and do not change as a
    function of $\tau$.  This is a special case of weight function that
    reduces dynamic correlations to static correlations.
    \textbf{b. Dirac $\delta$ kernel.} Only the observation at
    timepoint $t$ is given a non-zero weight (of 1).
    \textbf{c. Gaussian kernels.} Each kernel's weights fall off
    in time according to a Gaussian probability density function
    centered on time $t$.  Weights derived using several different
    example width parameters ($\sigma^2$) are displayed.
    \textbf{d. Laplace kernels.}  Each kernel's weights fall off
    in time according to a Laplace probability density function
    centered on time $t$.  Weights derived using several different
    example width parameters ($b$) are displayed.  \textbf{e. Mexican
      hat (Ricker wavelet) kernels.}  Each kernel's weights fall
    off in time according to a Ricker wavelet centered on time $t$.  This
    function highlights the \textit{contrasts} between local versus
    surrounding activity patterns in estimating dynamic
    correlations. Weights derived using several different example
    width parameters ($\sigma$) are displayed.}
  \label{fig:kernels}
\end{figure}

Given a kernel function $\kappa_t(\cdot)$ for timepoint $t$,
evaluated at timepoints $\tau \in \left[ 1, ..., T \right]$, we
can update the static correlation formula in Equation~\ref{eqn:corr}
to estimate the \textit{instantaneous correlation} at timepoint $t$:
\begin{align}
  \mathrm{timecorr}_{\kappa_t}\left(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)\right) &= \frac{\sum_{\tau=1}^T \left( \mathbf{X}(\tau, i) -
                                       \widetilde{\mathbf{X}}_{\kappa_t}(\cdot,
                                                                                        i) \right)
                                 \left( \mathbf{X}(\tau, j) -
                                        \widetilde{\mathbf{X}}_{\kappa_t}(\cdot,
                                                                                        j)\right)}
              {\sqrt{\sum_{\tau=1}^T
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot,
                                                                                        i))
                                              \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, j))}},~\mathrm{where}\label{eqn:timecorr}\\
  \widetilde{\mathbf{X}}_{\kappa_t}(t) &= \sum_{\tau=1}^T
                       \kappa_t(\tau, k)\mathbf{X}(\tau, k),\\
  \widetilde{\sigma}_{\kappa_t}^2(\mathbf{X}(\cdot, k)) &= \sum_{\tau=1}^T
                                                  \left(
                                                  \mathbf{X}(\tau, k) -
                            \widetilde{\mathbf{X}}_{\kappa_t}(t, k) \right)^2.
\end{align}
Here
$\mathrm{timecorr}_{\kappa_t}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot,
j))$ reflects the correlation at time $t$ between columns $i$ and $j$
of $\mathbf{X}$, estimated using the kernel $\kappa_t$.  We evaluate
Equation~\ref{eqn:timecorr} in turn each pair of columns in
$\mathbf{X}$ and for kernels centered on each timepoint in the
timeseries, respectively, to obtain a $T$ by $K$ by $K$ timeseries of
dynamic correlations, $\mathbf{Y}$.  For convenience, we then reshape
the upper triangles and diagonals of each timepoint's correlation matrix into a row
vector to obtain an equivalent $T$ by $\frac{K^2 - K}{2} + K$ matrix.

\subsubsection*{Dynamic inter-subject functional connectivity (DISFC)}
Equation~\ref{eqn:timecorr} provides a means of taking a single
observation matrix, $\mathbf{X}_n$ and estimating the dynamic
correlations from moment to moment, $\mathbf{Y}_{n+1}$.  Suppose that
one has access to a set of multiple observation matrices that reflect
the same phenomenon.  For example, one might collect neuroimaging data
from several experimental participants, as each participant performs
the same task (or sequence of tasks).  Let $\mathbf{X}_n^1$,
$\mathbf{X}_n^2$, ..., $\mathbf{X}_n^P$ reflect the $T$ by $K$
observation matrices ($n = 0$) or reduced correlation matrices
($n > 0$) for each of $P$ participants in an experiment.  We can use
\textit{inter-subject functional connectivity}~\citep[ISFC;
][]{SimoEtal16} to compute the degree of stimulus-driven correlations
reflected in the multi-participant dataset at a given timepoint $t$
using:
\begin{align}
\bar{\mathbf{C}}(t) = M\left(R\left(\frac{1}{2P} \sum_{p=1}^P
  Z\left(Y_n^p(t)\right)^\mathrm{T} + Z\left(Y_n^p(t)\right)\right)\right),\label{eqn:disfc}
\end{align}
where $M$ extracts and vectorizes the diagonal and upper triangle of a symmetric
matrix, $Z$ is the Fisher $z$-transformation~\citep{Zar10}:
\begin{align}
Z(r) = \frac{\log(1+r) - \log(1-r)}{2}
\end{align}
$R$ is the inverse of $Z$:
\begin{align}
R(z) = \frac{\exp(2z - 1)}{\exp(2z + 1)},
\end{align}
and $\mathbf{Y}_n^p(t)$ denotes the correlation matrix
(Eqn.~\ref{eqn:corr}) between each column of $\mathbf{X}_n^p$ and each
column of the average $\mathbf{X}_n$ from all \textit{other}
participants, $\bar{\mathbf{X}}_n^{ \backslash p}$:
\begin{align}
  \bar{\mathbf{X}}_n^{ \backslash p} = R\left(\frac{1}{P-1}\sum_{q \in
  \backslash p} Z\left( \mathbf{X}_n^q \right) \right),
\end{align}
where $\backslash p$ denotes the set of all participants other than
participant $p$. In this way, the $T$ by $\frac{K^2 - K}{2} + K$ DISFC
matrix $\bar{\mathbf{C}}$ provides a time-varying extension of the ISFC
approach developed by \cite{SimoEtal16}.

@@

Given a $T$ by $\frac{K^2 - K}{2} + K$ matrix of dynamic correlations,
$\mathbf{Y}_n$, we propose two general approaches to computing a $T$
by $K$ low-dimensional representation of these correlations,
$\mathbf{X}_n$.  The first approach uses dimensionality reduction
algorithms to project $\mathbf{Y}_n$ onto a $K$-dimensional space.
The second approach uses graph measures to characterize the
relative positions of each feature ($k \in \left[1, ..., K \right]$)
in the network defined by the correlation matrix at each timepoint.

@@

The modern toolkit of dimensionality reduction algorithms include
Principal Components Analysis~\citep[PCA; ][]{Pear01}, Probabilistic
PCA~\citep[PPCA; ][]{TippBish99}, Exploratory Factor
Analysis~\citep[EFA; ][]{Spea04}, Independent Components
Analysis~\citep[ICA; ][]{JuttHera91, ComoEtal91}, $t$-Stochastic
Neighbor Embedding~\citep[$t$-SNE; ][]{MaatHint08}, Uniform Manifold
Approximation and Projection~\citep[UMAP; ][]{McInHeal18},
non-negative matrix factorization~\citep[NMF; ][]{LeeSeun99},
Topographic Factor Analysis (TFA)~\cite{MannEtal14b}, Hierarchical
Topographic Factor analysis (HTFA)~\cite{MannEtal18}, Topographic
Latent Source Analysis~\cite[TLSA; ]{GersEtal11}, Dictionary
learning~\citep{MairEtal09a, MairEtal09b}, deep
auto-encoders~\citep{HintSala06}, among others.  While complete
characterizations of each of these algorithms is beyond the scope of
the present manuscript, the general intuition driving these approaches
is to compute the $T$ by $M$ matrix, $\mathbf{X}$, that is closest
to the original $T$ by $N$ matrix, $\mathbf{Y}$, where (typically)
$M \ll N$.  The different approaches place different constraints on
what properties $\mathbf{X}$ must satisfy and which aspects of
the data are compared (and how) to characterize the match between
$\mathbf{X}$ and $\mathbf{Y}$.

Applying dimensionality reduction algorithms to $\mathbf{Y}$ yields an
$\mathbf{X}$ whose columns reflect weighted combinations (or nonlinear
transformations) of the original columns of $\mathbf{Y}$.  This has
two main consequences.  First, with each repeated dimensionality
reduction, the resulting $\mathbf{X}_n$ has lower and lower fidelity
(with respect to what the ``true'' $\mathbf{Y}_n$ might have looked
like without using dimensionality reduction to maintain scalability).
In other words, computing $\mathbf{X}_n$ is a lossy operation.
Second, whereas each columns of $\mathbf{Y}_n$ may always be mapped
directly onto specific pairs of columns of $\mathbf{Y}_{n-1}$, the
columns of $\mathbf{X}_n$ reflect weighted combinations and/or
nonlinear transformations of the columns of $\mathbf{Y}_n$.  Many
dimensionality reduction algorithms are invertible (or approximately
invertible).  However, attempting to map a given $\mathbf{X}_n$ back
onto the original feature space of $\mathbf{Y}_0$ will usually require
$\mathcal{O}(TK^{2n})$ space and therefore quickly becomes intractable
as $n$ or $K$ grow large.
